---
title: "Searching Nanopubs"
output: html_notebook
---

# Searching on a Nanopublications server

The first step is to set up a search for the NanoPublications
REST API. We use the `curl` library for that, and the `jsonlite`
to process the JSON.

```{r}
library(curl)
library(jsonlite)
```

## Accessing a Nanopublication server

We also define where the REST API id found:

```{r}
searchAPI = "http://grlc.io/api/peta-pico/nanopub-api/find_latest_nanopubs_with_uri"
npServer = "http://graphdb.dumontierlab.com/repositories/nanopubs"
```

Let's say we are searching for ENSG00000181852:

```{r}
gene = "http://identifiers.org/ensembl/ENSG00000181852"
```

We define a handle with the information to make the API call. We want the results
returned as JSON and we use the `ref` parameter to pass the gene IRI:

```{r}
npHandle <- curl::new_handle()
curl::handle_setopt(npHandle, customRequest="GET")
curl::handle_setheaders(npHandle,
  "Accept" = "application/json"
)
```

With this search call defined, we can run it against the server, and retrieve
our JSON:

```{r}
queryURL = paste(searchAPI, "?ref=", curl_escape(gene),
  "&endpoint=", curl::curl_escape(npServer), sep=""
)
r <- curl::curl_fetch_memory(queryURL, npHandle)
jsonReply = rawToChar(r$content)
data = jsonlite::fromJSON(jsonReply)
```

We can count the number of returned nanopublications:

```{r}
nanopubs = data$results$bindings
if (length(nanopubs) == 0) nanopubs = as.vector(nanopubs)
if (length(nanopubs) == 1) nanopubs = as.vector(nanopubs[[1]][[2]])
cat(length(nanopubs))
```

## Equivalent IRIs with the BridgeDb Identifier Mapping Service

Let's define the location of our local service. This can be at your personal
computer or laptop, or a server on your local network. Instructions on how
to start a local IMS can be found [here](https://github.com/BiGCAT-UM/BridgeDbVariantDatabase/blob/master/Local_BridgeDb_IMS_installation.md).

```{r}
mapUriService = "http://localhost:8081/QueryExpander/mapUri"
# or a on a local network:
mapUriService = "http://192.168.0.113:8081/QueryExpander/mapUri"
```

But we cannot know IRI scheme or even from which database gene identifiers are used in
nanopublications. Therefore, we must use multiple IRIs for the same gene, and the BridgeDb
Identifier Mapping Service (IMS) gives those to us:

```{r}
h <- curl::new_handle()
curl::handle_setopt(h, customrequest = "POST")
curl::handle_setform(h, Uri=gene, format="application/json")
r <- curl::curl_fetch_memory(mapUriService, h)
jsonReply = rawToChar(r$content)
data = jsonlite::fromJSON(jsonReply)
mappings = length(data$Mapping$targetUri)
cat(paste("Mappings found:", mappings, "\n"))
data$Mapping$targetUri
```

## A convenience methods

For convenience, we define a function that will iterate over a list of IRIs to get nanopublications,
based on the earlier used code:

```{r}
getNanoPubs <- function(seedIRI,
  mapUriService = "http://192.168.0.113:8081/QueryExpander/mapUri",
  searchAPI = "http://grlc.io/api/peta-pico/nanopub-api/find_latest_nanopubs_with_uri",
  npServer = "http://graphdb.dumontierlab.com/repositories/nanopubs"
) {
  collectedNanopubs = c()
  
  # stuff for the identifier mapping
  h <- curl::new_handle()
  curl::handle_setopt(h, customrequest = "POST")
  curl::handle_setform(h, Uri=seedIRI, format="application/json")
  
  # stuff for the NP searching
  npHandle <- curl::new_handle()
  curl::handle_setopt(npHandle, customRequest="GET")
  curl::handle_setheaders(npHandle,
    "Accept" = "application/json"
  )

  # first, look up equivalent IRIs
  r <- curl::curl_fetch_memory(mapUriService, h)
  jsonReply = rawToChar(r$content)
  data = jsonlite::fromJSON(jsonReply)
  mappings = length(data$Mapping$targetUri)
  cat(paste("Mappings found:", mappings, "\n"))
  iris <- data$Mapping$targetUri

  # next, look up each mapped IRI
  directCount = 0
  for (iri in iris) {
    cat(paste("Processing:", iri, "... "))
    queryURL = paste(searchAPI, "?ref=", curl_escape(iri),
      "&endpoint=", curl::curl_escape(npServer), sep=""
    )
    r <- curl::curl_fetch_memory(queryURL, npHandle)

    jsonReply = rawToChar(r$content)
    data = jsonlite::fromJSON(jsonReply)
    nanopubs = data$results$bindings
    if (length(nanopubs) == 0) nanopubs = as.vector(nanopubs)
    if (length(nanopubs) == 1) nanopubs = as.vector(nanopubs[[1]][[2]])
    if (length(nanopubs) > 0) {
      cat(paste(" nanopubs:", length(nanopubs)))
      collectedNanopubs = c(collectedNanopubs, nanopubs)
      if (iri == seedIRI) {
        directCount = length(nanopubs)
      }
    }
    cat("\n")
  }
  return(
    list(
      seedIRI=seedIRI,
      mappings=iris,
      directCount=directCount,
      indirectCount=length(collectedNanopubs),
      nanopubs=collectedNanopubs
    )
  )
}
```

### Test

An example is given here:

```{r}
geneENSG000001658412 = getNanoPubs("http://identifiers.org/ensembl/ENSG00000165841")
```

# Finding nanopublications for WP241

First, list all gene IRIs from one of the popular pathways of WikiPathways:

```{r}
library(rrdf)

wikipathwaysEP = "http://sparql.wikipathways.org/"
geneQuery =  paste(
  "PREFIX wp:      <http://vocabularies.wikipathways.org/wp#>",
  "PREFIX xsd:     <http://www.w3.org/2001/XMLSchema#>",
  "PREFIX dcterms: <http://purl.org/dc/terms/>",
  "SELECT DISTINCT ?geneProduct WHERE {",
  "  ?geneProduct a wp:GeneProduct .",
  "  ?geneProduct dcterms:isPartOf/dcterms:identifier \"WP241\"^^xsd:string . ",
  "}"
)

pathwayGenes = rrdf::sparql.remote(
  wikipathwaysEP, geneQuery
)
```

This returns about 28 unique gene IRIs, all using an IRI pattern for NCBI Gene identifier,
e.g. http://identifiers.org/ncbigene/1719. We can iterate over them, and for each look up
nanopublications:

```{r}
results = lapply(pathwayGenes, getNanoPubs)
```

From these resulst we can see the number of additional nanopublications we get back,
starting from the initial gene IRI:

```{r}
counts = matrix(
  data=unlist(
    lapply(results, function(x) { c(x$directCount, x$indirectCount) } )
  ),
  ncol=2, byrow=TRUE
)
colnames(counts) = c("direct", "indirect")
extraGenes = counts[,"indirect"]-counts[,"direct"]
```

Similarly, we can also take a pathways with predominantly Ensembl identifiers:

```{r}
wikipathwaysEP = "http://sparql.wikipathways.org/"
geneQuery =  paste(
  "PREFIX wp:      <http://vocabularies.wikipathways.org/wp#>",
  "PREFIX xsd:     <http://www.w3.org/2001/XMLSchema#>",
  "PREFIX dcterms: <http://purl.org/dc/terms/>",
  "SELECT DISTINCT ?geneProduct WHERE {",
  "  ?geneProduct a wp:GeneProduct .",
  "  ?geneProduct dcterms:isPartOf/dcterms:identifier \"WP2059\"^^xsd:string . ",
  "}"
)

pathwayGenes = rrdf::sparql.remote(
  wikipathwaysEP, geneQuery
)
```

```{r}
results2 = lapply(pathwayGenes, getNanoPubs)
```

```{r}
counts2 = matrix(
  data=unlist(
    lapply(results2, function(x) { c(x$directCount, x$indirectCount) } )
  ),
  ncol=2, byrow=TRUE
)
colnames(counts2) = c("direct", "indirect")
extraGenes2 = counts2[,"indirect"]-counts2[,"direct"]
```

We now report some statistical characterizations of the results for the two pathways.

First, we want to know how many equivalent IRIs were returned by the IMS:

```{r}
mappings = unlist(
  lapply(results, function(x) { length(x$mappings) } )
)
summary(mappings)
mappings2 = unlist(
  lapply(results2, function(x) { length(x$mappings) } )
)
summary(mappings2)
```

Statistics for the direct hits for the two pathways are as follows:

```{r}
summary(counts[,"direct"])
summary(counts2[,"direct"])
```

Statistics for the all hits for the two pathways after IRI mapping are as follows:

```{r}
summary(counts[,"indirect"])
summary(counts2[,"indirect"])
```
